Determining the direction of causality in the face of measurement error

### 2016-02-14 docs/manuscript.tex
updated

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{22}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\). To infer the direction of
causality using the CIT method, the omnibus p-value was estimated for
\(x\) causing \(y\), and for \(y\) causing \(x\), and the model that
returned the lowest p-value was taken as the one denoting the correct
causal direction.

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a
population\textsuperscript{\textbf{???}}. It is calculated as

@@

\[
\begin{aligned}
cor(g, x) & = \{0.01, 0.05, 0.1\} \\
cor(x, y) & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables.

@@

In the epidemiological literature this is relatively well understood and
our initial analysis simply confirms that for widely used methods such
as CIT that it is indeed liable to the same issues as standard mediation
based analysis. Under many circumstances a practical solution to this
problem is to use Mendelian randomisation instead of mediation based
methods. Mendelian randomisation is robust in the face of measurement
error and, given that the researcher has knowledge about the biology of
the instrument being used in the analysis, can offer a direct solution
to the issues that CIT faces. This assumption is often reasonable, for
example exposures such as CRP

Pleiotropy

@@

\[
x = \alpha_g + \beta_g g + \epsilon_g \\
x_O = \alpha_m + \beta_m x + \epsilon_m \\
y = \alpha_x + \beta_x x + \epsilon_x
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = D\beta_x\beta_g var(g)
\end{aligned}
\]

\[
D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\]

@@

\subsection{Tables}\label{tables}

\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}\\
Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations.

\subsection{Supplementary tables}\label{supplementary-tables}

\subsection{Supplementary figures}\label{supplementary-figures}

@@

22.Millstein, J. \emph{cit: Causal Inference Test}. (2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

>>>>

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{22} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{23}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\). To infer the direction of
causality using the CIT method, the omnibus p-value was estimated for
\(x\) causing \(y\), and for \(y\) causing \(x\), and the model that
returned the lowest p-value was taken as the one denoting the correct
causal direction.

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population\textsuperscript{24}. It
is calculated as

@@

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables. Similar patterns of results were obtained for
different values of \(cor(g, x)\).

@@

In the epidemiological literature this is relatively well understood and
our initial analysis simply confirms that for widely used methods such
as CIT they are indeed liable to the same issues as standard mediation
based analysis. Specifically, we show that as measurement error in the
exposure variable increases, CIT is likely to infer the wrong direction
of causality. We also demonstrate that, though seemingly unintuitive,
increasing sample size does not resolve the issue, rather it increases
the significance of the incorrect inference.

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of mediation based methods. Mendelian
randomisation is robust in the face of measurement error and, given that
the researcher has knowledge about the biology of the instrument being
used in the analysis, can offer a direct solution to the issues that CIT
faces. This assumption is often reasonable, for example SNPs are
commonly used as instruments when they are found in genes with known
biological relevance to the trait of interest. But on many occasions
this is not the case, and it may be tempting to resort to using methods
based on mediation in order to be able to both ascertain if there is a
causal association and to infer the direction of causality. Here we have
described a simple extension to MR which can be used as an alternative
to mediation based methods. We show that this method is still liable to
measurement error, but because it has different properties to the CIT it
offers two main advantages. First, it uses a formal statistical
framework to test for the robostness of the assumed direction of
causality. Second, after testing in a comprehensive range of scenarios
the MR based approach is much less likely to infer the wrong direction
of causality compared to CIT.

There are a number of limitations to the analysis that we performed. We
focused on the simple case of a single instrument in a single sample
setting, and assumed that pleiotropy (the influence of the instrument on
the outcome through a mechanism other than the exposure) was not
present.

The overarching result from our simulations is that, regardless of the
method used, inferring causal direction using an instrument of unknown
biology is very susceptible to measurement error. With the presence of
measurement error near ubiquitous in most cross sectional studies, and
our ability to measure it limited, we argue that it should be given
serious consideration in designing statistical analyses that attempt to
make causal inference, and that any putitive results should be
accompanied with appropriate sensitivity analysis that assesses their
robustness under varying levels of measurement error.

Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations.

@@

\[
\begin{aligned}
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y & = \alpha_x + \beta_x x + \epsilon_x
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
\end{aligned}
\]

@@

@@

22.R Core Team. \emph{R: A Language and Environment for Statistical
Computing}. (R Foundation for Statistical Computing, 2015). at
\textless{}\url{https://www.r-project.org/}\textgreater{}

23.Millstein, J. \emph{cit: Causal Inference Test. R package version
1.9}. (2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

24.Steiger, J. H. Tests for comparing elements of a correlation matrix.
\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
